Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-18 src/tex/ms.tex
Finishing result section

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their intrinsic
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of optimized waveforms with original waveforms. We also show how the
optimization result differs with waveforms of different intrinsic parameters. In
Sec.~\ref{sec:discussion}, we address the difference between our calibrating
procedure with \citep{khan2016frequency}. We also explain how reduced spin
parameterization affects the accuracy of the model.

@@

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248}. We 
specifically choose to use $q=2$ events since we have abundant $q=2$ NR waveforms 
to validate the final result. The new set of coefficients generated from this
optimization process yields only marginal improvements in the newly produced
waveforms, as seen in Fig. \ref{fig:q148_q1248_compare}. The high mismatch tail
of the optimized distribution remains comparable in length and endpoint to the
original distribution, indicating that our procedure is incapable of improving
these waveforms. Similarly, utilizing the {\zdethp} PSD to optimize the loss
function with additional waveforms leads to similar improvement in the resulting
distribution.

% We believe that either the ansatz does not suit waveforms within certain
% regions of parameter space or the set of optimized coefficients falls to a
% minimum where they do not fit some testing waveforms. If in some region of
% parameter space, the ansatz does not fit well with NR waveforms, it would
% never be able to have consistent improvement under optimization. Instead, it
% would surf around the loss manifold with random fluctuations in mismatches. If
% the set of coefficients does not fit, then dividing the parameter space into
% regions and fitting separate sets of coefficients should give better results.
Although most of the waveforms show improvement in figure \ref{fig:q148}, the
high mismatch tail of the distribution remains unaffected. Given that the
waveform model's ansatz may not be entirely compatible with NR, and the
optimization procedure is carried out over a distribution of waveforms with
varying intrinsic parameters, it is conceivable that some trade-offs in accuracy
exist between different parts of the parameter space. If this is the cause of
the lack of improvement in the high mismatch tail of the distribution,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right ($\chi_1>0>\chi_2$) 
regions respond to optimization differently. In the top-left 
region, waveforms generally improve slightly with along optimization. However, waveforms 
in the bottom-right region do not improve after optimization. Some waveforms even 
turned worse after optimization.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148_quadrant}). 
The training waveforms used for fitting are listed in Tab.~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\lesssim4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe that 
equal-spin waveforms lying on the diagonal has great improvement. Above the diagonal, 
the improvement is significant, except for a few defects caused by some testing waveforms 
having $q>4$. This can be seen clearer in Fig.~\ref{fig:all_quadrants}, which we see 
that the optimized histogram shifts downward uniformly. Below the diagonal, waveforms 
have negligible improvement (Fig.~\ref{fig:all_quadrants}~and~\ref{fig:ps_q148_quadrant}), 
indicating the original set of coefficients is the optimal set of coefficients, and 
cannot be further improved. Hence, we can deduce IMRPhenomD has better match with the 
top-left region and does not fit well with waveforms lying in the bottom-right 
region. Thus, the ansatz restricts additional improvements and further optimizing in a 
smaller region does not generally improve the results.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left (Top) and bottom-right (Bottom) regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ as the loss function. Waveforms in the top-left region generally improves while waveforms in the bottom-right region has very little improvement, as indicated by the median (dashed lines).}
	\label{fig:all_quadrants}
\end{figure}

@@

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region hardly improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region only have little 
improvement. This indicates that the ansatz has limited match in this region and 
does not fit most waveforms in this region, hence introduces bias to downstream GW 
analysis. This phenomenon is mainly due to the reduced spin approximation used in 
parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.

>>>>

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their source
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.

The rest of the paper is structured as follows: In Sec.~\ref{sec:method}, we
review the parameterization of the IMRPhenomD model and the mismatch function
that is used as an objective function for the calibration, followed by outlining the
specific optimization scheme used for recalibration. In
Sec.~\ref{sec:result}, we give the optimization result by comparing mismatches
of optimized waveforms with original waveforms. We also show how the
optimization result differs with waveforms of different source parameters. In
Sec.~\ref{sec:discussion}, we address the difference between our calibrating
procedure with \citep{khan2016frequency}. We also explain how reduced spin
parameterization affects the accuracy of the model.

@@

To understand whether additional training data can further improve the
performance of the model, we include waveforms that are not presence in the
original IMRPhenomD calibration in our training dataset with parameters
tabulated in Tab. \ref{tab:q1248}. We specifically choose to use $q=2$ events
since we have abundant $q=2$ NR waveforms to validate the final result. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the optimized
distribution remains comparable to the original distribution, indicating that
the original dataset is sufficient for this task. Similarly, utilizing the
{\zdethp} PSD to optimize the loss function with additional waveforms results in 
similar level of improvement.

To investigate the performance of recalibration over the source parameter space,
we plot the improvement of mismatch in log projected over the parameter space of
$q$ vs. $\chi_{\mathrm{PN}}$ in Fig. \ref{fig:ps_q148_qchi}. On the mass ratio
axis, we can see the waveforms with $q\lq4$ show most consistent average
improvement. This is because that part of the parameter space is better covered
by the training waveform, as compared to testing waveforms from $q=4$ to $q=8$,
where some testing waveforms lie outside the source parameter space covered by
the training set of waveforms. On the spin axis, we can see the waveform with
$\chi_{\textrm{PN}}$ close to zero show the most consistent improvement, and we
start deviate from 0, there could be fluctuations in the average improvement. On
top of that, in the $q\lq4$ region, there seems to be more consistent
improvement of waveforms with $\chi_{\textrm{PN}}<0$. This can be understood as
since the original waveform was developed for aligned spin system, the waveforms
with $\chi_{\textrm{PN}}$ is less well-fitted \kw{Varify this.}, hence there
could be more room for improvement. To show this, we plot the parameter space of
$\chi_1$ vs. $\chi_2$ in Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal
axis, i.e. $\chi_1\approx\chi_2$, show good mismatch improvements as discussed
above. Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right
($\chi_1>0>\chi_2$) regions respond to optimization differently. In the top-left
region, waveforms generally improve slightly with along optimization. However,
waveforms in the bottom-right region do not improve after optimization. Some
waveforms even turned worse after optimization.

 Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying source parameters, it is conceivable that different parts of the
source parameter space may not share the same set of optimal IMRPhenomD
parameters, mean there could be some trade-offs in accuracy between different
parts of the parameter space. If this is the cause of the lack of improvement in
the high mismatch tail of the distribution, segmenting the parameter space into
smaller subspaces should alleviate this problem. On the other hand, if the
ansatz lacks the correct parameterized form to capture the NR waveforms'
behavior as a function of the source parameters, the results will always be
biased, and we should not expect any improvement, even if we segment the
parameter space during training.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants}. The top-left and
bottom-right regions have limited data for $q>4$, hence the results are only
valid up to $q\lesssim4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe that
equal-spin waveforms lying on the diagonal has great improvement. The
improvement above the diagonal is also substantial compared to recalibrate using
all waveforms at once, barring a few caused by some testing waveforms having
$q>4$. This can be seen clearer in Fig.~\ref{fig:all_quadrants}, which we see
that the optimized histogram shifts downward uniformly. On the other hand, the
improvement below the diagonal is negligible, indicating our method struggles to
further improve on top of the original IMRPhenomD waveform. Interestingly, we
can see in the top-right and bottom-left quadrant, the lower right corner shows
systematic worsening similar to the case of all waveforms. This hinted in
general, the IMRPhenomD waveform ansatz does not fit well with waveforms with
anti-aligned spin where the primary spin is positive.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left (Top) and bottom-right
	(Bottom) regions. We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ as the loss function. Waveforms in the
	top-left region generally improves while waveforms in the bottom-right
	region has very little improvement, as indicated by the median (dashed
	lines). \kw{Is the original referring to original IMRPhenomD or optimized
	using everything? Can we have a reference for optimizing using everything?}}
	\label{fig:all_quadrants}
\end{figure}

@@

Furthermore, we investigated how the source parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region hardly improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region only have little 
improvement. This indicates that the ansatz has limited match in this region and 
does not fit most waveforms in this region, hence introduces bias to downstream GW 
analysis. This phenomenon is mainly due to the reduced spin approximation used in 
parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.
